%-----------------------
%-----------------------
\chapter*{Acknowledgements}
%-----------------------
%-----------------------
This thesis would not have been completed without the help of others. I would like to take this opportunity to express my gratitude towards them and acknowledge them.

First of all, I would like to offer my deepest gratitude to my supervisors \textit{Olivier Barais} and \textit{Benoit Baudry} for their useful advices, and for guiding my steps over more than three years. Special thanks are also due to my ``third" supervisor \textit{Gesron Suny\'e} that always knew how to motivate me and help me to overcome the difficult times in my thesis. 

I owe special thanks to my Master's supervisor \textit{Marouane Kessentini}, with whom I discovered the software engineering research world, for encouraging me to continue in the research field as a PhD candidate.

I would also like to thank the members of my jury. My examiners, \textit{H\'el\`ene Waeselynck} and \textit{Philippe Merle} for having accepted to serve on my examination board and for the time they have invested in reading and evaluating this thesis. Thanks are also due to \textit{Erven Rohou} for agreeing to preside over my jury. I would also like to thank \textit{Jean-Marie Mottou} and \textit{Franck Fleury} for attending as members of the jury. 
%You all gave me a lot of interesting and inspiring feedback that I will keep in mind in the future.

I greatly appreciate and wish to thank all the (former and current) members of the DiverSE research team for providing a great and friendly work atmosphere, and for all the different discussions, seminars, and DiverSE coffees. 

Finally, I would like to thank my family. Even so far away from home, they are always present in my life during these three years. I first express my gratitude to my parents \textit{Najoua} and \textit{Noureddine} for their support, comprehension, confidence and, especially for their patience. You have always been behind me and believed in me. Then, thanks to my sisters \textit{Imene} and \textit{Jihene} for encouraging me to succeed in everything I do. 
By now, we can say that we have a Doctor in the \textit{Boussaa} family !

